---
title: "Data analysis: Meta-analysis of the influence of ENSO phases on yield-loss caused by soybean rust in Brazil"
author: 
 - name: Kaique S. Alves
   affiliation: "Universidade Federal de Viçosa, Viçosa MG, Brazil"
date: last-modified
title-block-banner: true
format:
  html:
    code-fold: false
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 5
    # df-print: paged
    self-contained: true
    anchor-sections: true
    smooth-scroll: true
    code-copy: true
---


# Packages

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(gsheet)
library(cowplot)
library(patchwork)
library(lemon)
library(lme4)
library(ggforce)
library(lmerTest)
library(emmeans)
library(multcomp)
library(ggthemes)
library(metafor)
library(minpack.lm)
```

```{r}
theme_set(theme_half_open())
```


# Data

## Soybean Rust data

```{r}
data_load = gsheet2tbl("https://docs.google.com/spreadsheets/d/1eGcLU8JVJ485pkKgAQUuVorapp57TVvYtRO5hpnmW3w/edit#gid=1365448637")
```

## ENSO data
```{r}
enso_data = gsheet2tbl("https://docs.google.com/spreadsheets/d/1NFm1hKCS8V4r3GRbE13QbKAQ8RZ0e21WES6UtN7uM_8/edit?usp=sharing")
```

Classifying the years based on Oceanic Niño Index (ONI) on the October, November, and December (OND) trimester. Seasons with ONI higher then 75 percentiles were classified as warm, year with ONI lower then its 25 percentiles were classified as Cold, and the years with ONI within the 25 and 75 percentiles were classified as neutral.

```{r}
enso_data_class = enso_data %>%
  mutate(year = as.character(Year+1)) %>% 
  dplyr::select(-Year) %>% 
  dplyr::select(OND, year) %>%
  filter(!is.na(OND),
         year != 2021) %>% 
  mutate(enso = case_when(OND > quantile(OND, 0.75) ~ "Warm",
                          OND < quantile(OND, 0.25) ~ "Cold",
                          OND < quantile(OND, 0.75) & OND > quantile(OND, 0.25) ~ "Neutral"),
         year = as.numeric(year))
enso_data_class
```

```{r}
enso_data %>%
  mutate(year = as.character(Year+1)) %>% 
  dplyr::select(-Year) %>% 
  dplyr::select(OND, year) %>%
  filter(!is.na(OND),
         year != 2021) %>% 
  summarise(low = quantile(OND, 0.25),
            high =  quantile(OND, 0.75) )
```


```{r}
enso_data_class%>% 
  mutate(enso = factor(enso, levels = c("Neutral", "Warm", "Cold")))%>% 
  ggplot(aes( as.factor(year),OND,  color = enso))+
  geom_hline(yintercept = 0)+
  geom_hline(yintercept = c(-0.725,0.75), linetype = 2, color = "gray")+
  geom_errorbar(aes(ymin=0, ymax  = OND), width = 0, color = "black")+
  geom_point(size = 3)+
  
  scale_color_colorblind()+
  theme_half_open(font_size = 12)+
  # background_grid()+
  scale_y_continuous(breaks = c(-0.725,0.75), limits = c(-2,2.6))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        # legend.position = "top",
        strip.background = element_blank())+
  labs(x = "Crop Season",
       y = "Oceanic Niño Index",
       color = "ENSO phase (OND)")
ggsave("figs/ONI_OND.png", dpi = 600, height = 4, width = 7, bg = "white")
```

Counting the number of year for each phase
```{r}
enso_data_class %>% 
  count(enso)
```


# Data wrangling


```{r}
data_load2 = data_load %>%
  # mutate(enso = case_when(year %in% c(2007,2010,2016,2019) ~ "EL Nino",
  #                         year %in% c(2008,2011, 2012,2021) ~ "La Nina",
  #                         !year %in% c(2007,2010,2016,2019,2008,2011, 2012,2021)~"Neutral"),
  #        enso = factor(enso, levels = c("Neutral", "EL Nino", "La Nina"))
  # ) %>% 
  full_join(enso_data_class) %>% 
  mutate(enso = factor(enso, levels = c("Neutral", "Warm", "Cold"))) %>% 
  filter(!is.na(sev),
         !is.na(yld)) %>% 
  mutate(study = as.factor(study)) %>% 
  mutate(region = case_when(state %in% c("SP","BA","MG", "MS", "MT", "GO",  "MA", "DF", "TO")~"North",
                             state %in% c("RS","SC","PR") ~"South"),
         region =factor(region, levels = c("South","North"))) %>% 
  group_by(study) %>% 
  mutate(difer = max(sev) - min(sev)) %>% 
  filter(difer>5)



data_load2 %>% 
  filter(active_ingred == "CHECK") %>% 
  ggplot(aes(sev, difer))+
  geom_point()

```

```{r}
unique(data_load2$state)
length(unique(data_load2$state))

```

```{r}
data_load2 %>% 
  group_by(state) %>% 
  summarise(n_loc = length(unique(location))) %>% 
  summarise(sum(n_loc))
```

```{r}
length(unique(data_load2$year))
```


## Tranformations

Converting percent severity into proportion and calculating logits

```{r fig.height=4, fig.width=12}
only_check_df = data_load2 %>% 
  filter(active_ingred == "CHECK") %>% 
  mutate(sev = case_when(sev == 100 ~ 99.9,
                         sev == 0 ~ 0.1,
                         sev>0 & sev<100 ~ sev), 
           logit_sev =  DescTools::Logit(sev/100))
```

# Modeling Severity (untreated) and ENSO



```{r}
enso_sev_gg= data_load2 %>% 
  filter(active_ingred == "CHECK") %>% 
  ggplot(aes(enso, sev, color = enso))+
  geom_sina(color = "gray")+
  geom_boxplot(fill= NA, size = 1)+
  theme_half_open(font_size = 12)+
  scale_color_colorblind()+
  labs( y = "",
        x = "",
        color = "",
        title = "")
enso_sev_gg
```

## Graph over time


```{r fig.height=4, fig.width=12}
sev_untreat_gg = only_check_df %>% 
  ggplot(aes(as.factor(year),  sev, color = enso))+
  geom_sina(color = "gray80")+
  geom_boxplot(fill =NA, size = 1, outlier.color = NA)+
  labs( y = "Severity (%)",
        x = "",
        color = "",
        title = " ")+
  theme_half_open(font_size = 12)+
  # facet_wrap(~region)+
  scale_color_colorblind()+
  theme(
    # axis.text.x = element_text(angle = 25, hjust = 1)
    )
sev_untreat_gg
# ggsave("figs/severity_check.png", dpi = 600, height = 4, width = 10, bg = "white")
```


## Mixed-effect model

```{r}
model_check = lmer(logit_sev ~ enso+ (1|year/study), data = only_check_df,REML = F)
```

### Model summary

```{r}
summary(model_check)
```

### Pairwise comparison

```{r}
pairs(emmeans(model_check, ~enso))
```

### All Severity data (plot)

```{r fig.height=4, fig.width=12}
sev_gg = data_load2 %>% 
  ggplot(aes(as.factor(year),  sev, color = enso))+
  geom_sina(color = "gray80")+
  geom_boxplot(fill =NA, size = 1, outlier.color = NA)+
  labs( y = "Severity (%)",
        x = "Growing season",
        color = "",
        title = "Data from all plots")+
  theme_half_open(font_size = 12)+
  # facet_wrap(~region)+
  scale_color_colorblind()+
  theme(
    # axis.text.x = element_text(angle = 25, hjust = 1)
    )
sev_gg

```


# Modeling soybean yield and ENSO


```{r}
enso_yld_gg = data_load2 %>% 
  filter(active_ingred == "CHECK") %>% 
  ggplot(aes(enso, yld, color = enso))+
  geom_sina(color = "gray")+
  geom_boxplot(fill= NA, size = 1)+
  scale_color_colorblind()+
  theme_half_open(font_size = 12)+
  labs( y = "",
        x = "",
        color = "",
        title = "")
enso_yld_gg
```


## Graph over time

```{r fig.height=4, fig.width=12}

yld_untreat_gg = data_load2 %>% 
  filter(active_ingred == "CHECK")%>% 
  ggplot(aes(as.factor(year),  yld, color = enso))+
  geom_sina(color = "gray80")+
  geom_boxplot(size =1, fill = NA, outlier.colour = NA)+
  labs( y = "Yield (kg/ha)",
        x = "Growing season",
        color = "",
        title = "")+
  theme_half_open(font_size = 12)+
  scale_color_colorblind()+
  theme(
    # axis.text.x = element_text(angle = 25, hjust = 1)
    )
yld_untreat_gg

```


## Mixed-effect model

```{r}
model_check_yld = lmer(yld ~ enso + (1|year/study), data = only_check_df, REML = F)
```

### Model summary

```{r}
summary(model_check_yld)
```

### Pairwise comparison

```{r}
cld(emmeans(model_check_yld, ~enso))
```

### All yield data

## Graph over time

```{r fig.height=4, fig.width=12}

yld_gg = data_load2 %>% 
  ggplot(aes(as.factor(year),  yld, color = enso))+
  geom_sina(color = "gray80")+
  geom_boxplot(size =1, fill = NA, outlier.colour = NA)+
  labs( y = "Yield (kg/ha)",
        x = "Growing season",
        color = "",
        title = "Data from all plots")+
  theme_half_open(font_size = 12)+
  scale_color_colorblind()+
  theme(
    # axis.text.x = element_text(angle = 25, hjust = 1)
    )
yld_gg

```






## Data combo

```{r}
sev_untreat_gg+ enso_sev_gg+
  yld_untreat_gg +enso_yld_gg+
  plot_annotation(tag_levels = "A")+
  plot_layout(ncol = 2,
              widths = c(1,0.25),
              guides = "collect")&
  theme(axis.text = element_text(size =8))

ggsave("figs/data_over_time.png", dpi = 600, height = 6, width = 9, bg = "white")
```



# Modeling disease damage

## Yield vs. Severity

```{r}
data_load2 %>% 
  ggplot(aes(sev, yld, color = enso))+
  # geom_point(alpha = 0.1)+
  geom_smooth(aes(group=study),
              method = "lm", size =0.1,se = F, fullrange   =T)+
  geom_smooth(aes(group = enso),  color = "red", se = F, method = "lm")+
  scale_color_colorblind()+
  theme_half_open()+
  facet_rep_grid(~enso)+
  ylim(0,7500)+
  labs(x = "Severity (%)",
       y = "Yield (kg/ha)",
       color = "")+
  theme(strip.background = element_blank())

```




## Meta-analysis

###  Ordinary Regression 

```{r}
reg_dc = data_load2 %>% 
  group_by(study, year, region, enso) %>% 
  summarise(intercept = lm(yld~sev)$coefficients[1],
            slope = lm(yld~sev)$coefficients[2],
            r2 = summary(lm(yld~sev))$r.squared,
            sigma = summary(lm(yld~sev))$sigma) %>%
  mutate(Dc = (slope/intercept)*100) %>% 
  filter(Dc<0.5)
reg_dc



```


#### Graph of the regression lines

```{r}
reg1_gg = reg_dc %>% 
  ggplot()+
  geom_abline(aes(intercept = intercept, slope = slope, color= enso), alpha = 0.9)+
  scale_color_colorblind()+
  labs(x = "Severity (%)",
       y = "Yield (kg/ha)")+
  theme_half_open()+
  xlim(0,100)+
  ylim(0,7500)+
  scale_color_colorblind()+
  theme_half_open(font_size = 10)+
  facet_rep_grid(~enso)+
  labs(x = "SBR severity (%)",
       y = "Soybean yield (kg/ha)",
       color = "")+
  theme(strip.background = element_blank(),
        legend.position = "none")
reg1_gg
```




#### Distribution of intercepts and slopes

```{r}

reg_dc %>% 
  ggplot(aes(intercept))+
reg_dc %>% 
  ggplot(aes(slope))+
   
#patchwork
plot_layout(nrow = 1)+
  plot_annotation(tag_levels = "A")&
  
#more ggplot  
  geom_histogram(color = "white", bins = 10)&
  theme_half_open()
```




### Percent relative yield loss relative to β~0~

```{r}
data2 = data_load2 %>% 
  full_join(reg_dc) %>% 
  mutate(l = 100*((intercept - yld)/intercept)) %>% 
  filter(!is.na(l))

data2
```



```{r}
data2 %>% 
  ggplot(aes(sev,l))+
  geom_point(color = "gray90")+
  geom_smooth(method ="lm", aes(color =enso), formula = y~0+x)+
  scale_color_colorblind()+
  labs(x = "Severity (%)",
       y = expression("Yield loss relative to "~β[0]))+
  theme_half_open()
```


#### Second Regression l ~ 0 + sev

```{r}
data3 = data2 %>% 
  group_by(study, year, enso, region) %>% 
  summarise(slope = lm(l~0+sev)$coefficients,
            vi = as.numeric(vcov(lm(l~0+sev)))
            ) %>% 
  full_join(enso_data_class)
data3
```


```{r}
reg2_gg =
  data3 %>% 
  ggplot()+
  geom_abline(aes(intercept = 0, slope = slope, color= enso), alpha = 0.9)+
  scale_color_colorblind()+
  labs(x = "Severity (%)",
       y = "Yield (kg/ha)")+
  theme_half_open()+
  xlim(0,100)+
  ylim(0,100)+
  scale_x_continuous(expand = c(0, 0),
                     limits = c(0,100)) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0,100))+
  scale_color_colorblind()+
  theme_half_open(font_size = 10)+
  facet_rep_grid(~enso)+
  labs(x = "SBR severity (%)",
       y = expression("Yield loss relative to "~β[0]~"(%)"),
       color = "")+
  theme(strip.background = element_blank(),
        legend.position = "none")
reg2_gg
```







#### Damage vs. ENSO phases

```{r}
data3 %>% 
  ggplot(aes(slope))+
  geom_histogram(color = "gray")+
  facet_rep_wrap(~enso, ncol=1 )
```



```{r}
library(scales)
slopes_gg = data3 %>% 
  arrange(slope) %>% 
  ggplot(aes(enso, slope ))+
  geom_boxplot(fill = NA, color = "steelblue", size =1, outlier.colour = NA)+
  geom_sina(aes(size = vi, color= vi), alpha = 0.5)+
  scale_size_continuous(range = c(1, 7),
                        #limits =  c(0.01,1),
                        breaks = c(0.001, 0.01, 0.1, 1),
                       label=scientific_format())+
  scale_color_gradient(low = "black",
                       high = "red",
                       #limits = c(0.01, 1), 
                        breaks = c(0.001, 0.01, 0.1, 1),
                       label=scientific_format())+
  guides(color= guide_legend(), 
         size=guide_legend())+
  theme_half_open(font_size = 10)+
  labs(y = "Slope",
       x = "",
       size = "Slope variance",
       color  = "Slope variance" )
slopes_gg

```

```{r}
(reg1_gg+reg2_gg)+
  plot_annotation(tag_levels = "A")+
  plot_layout(ncol = 1)

ggsave("figs/regression_lines1.png", dpi = 900, width = 8, height = 5, bg ="white")
```


```{r fig.height=5, fig.width=8}
layout <- "
AAAACC
BBBBCC
"

(reg1_gg+reg2_gg+slopes_gg)+
  plot_annotation(tag_levels = "A")+
  plot_layout(design = layout)

ggsave("figs/regression_lines.png", dpi = 600, width = 8, height = 5, bg ="white")
```


### Meta-analytical model

```{r}
data4 = data3 %>% mutate(year = as.factor(year))
metamodel2 = rma.uni(yi = slope,
                    vi = vi,
                    mods = ~ 0+enso, 
                    random = list(~1|year/study),
                    # struct = "HCS",
                    method = "ML",
                    data =data4)
metamodel2

```



#### Estimates for each moderator

```{r}


grid = qdrg(object = metamodel2,  data = data4, at = list(vi = 0, year.c = 0))
# cld(emmeans(grid,specs = ~enso, by = "region"), Letters = letters)
cld(emmeans(grid, specs = ~enso), Letters = letters)

# grid@V
```

### Graphs

#### Forest plot

```{r}
forest_gg = data3 %>% 
  mutate(study = as.character(study),
         sd = sqrt(vi),
         
         cil = slope - 1.96*sd,
         ciu = slope + 1.96*sd) %>% 
  full_join(as.data.frame(emmeans(grid,specs = ~enso))) %>% 
  mutate(enso = factor(enso, levels = c("Neutral",  "Warm","Cold"))) %>% 
  ggplot(aes(slope,reorder(study, slope),  color= enso))+
  geom_point(size=0.3)+
  geom_errorbar(aes(xmin=cil, xmax = ciu), width = 0,size=0.3, alpha =0.5)+
  geom_vline(xintercept = 0, linetype = "dashed")+
  geom_vline(aes(xintercept = emmean), color = "gray40", size = .5)+
  scale_color_colorblind()+
  facet_rep_grid(rows = vars(enso),  scales="free_y")+
  labs(y = "",
       x = "Damage coefficient (%/p.p.)",
       color = "")+
  guides(color = guide_legend(override.aes = list(size=2.5)))+
  theme_half_open()+
  theme(axis.text.y = element_blank(),
        axis.ticks.length.y = unit(0, "cm"),
        strip.background = element_blank(),
        legend.position = "right")
forest_gg

# ggsave("figs/z_studies.png", dpi = 600, width = 5, height = 8, bg ="white")
```


#### Damage coefficient

```{r}
dc_data =  as.data.frame(emmeans(grid, specs = ~enso))
dc_data
```

```{r}

DC_gg = as.data.frame(emmeans(grid,specs = ~enso)) %>% 
    mutate(enso = factor(enso, levels = c("Neutral",  "Warm","Cold"))) %>% 
  ggplot(aes(reorder(enso,emmean), emmean, color = enso))+
  geom_point(position = position_dodge(width = 0.2), size= 3)+
  geom_errorbar(aes(ymin =asymp.LCL, ymax = asymp.UCL),
                position = position_dodge(width = 0.2),
                width = 0,
                size = 1)+
  labs(x = "ENSO phases",
       y = "Damage coefficient (%/p.p.)",
       color = "")+
  theme_half_open()+
  # background_grid()+
  scale_y_continuous(breaks = seq(0.1,0.8, by = 0.1), limits =c(0.4,0.8))+
  scale_color_colorblind()+
  theme(legend.position = "none")
  
 
DC_gg
```


#### Yield loss

```{r}


rel_gg = as.data.frame(emmeans(grid,specs = ~enso)) %>% 
    mutate(enso = factor(enso, levels = c("Neutral",  "Warm","Cold"))) %>% 
  mutate(`100` = 100-100*emmean,
         `0` = 100) %>% 
  pivot_longer(7:8,
               names_to = "sev", 
               values_to = "yloss") %>%
  mutate(sev = as.numeric(sev)) %>%
  mutate(cil = 100-sev*asymp.LCL,
         ciu = 100-sev*asymp.UCL) %>% 
  mutate(yl = -(yloss-100)) %>% 
  # mutate(cil = ifelse(cil==100, cil, -cil),
  #        ciu = ifelse(ciu==100, ciu, -ciu)) %>% 
  ggplot(aes(sev, yloss,color =  enso, fill = enso ))+
  geom_ribbon(aes(ymin = cil, ymax = ciu),alpha = 0.5, color = NA)+
  geom_line(size = 1)+
  # geom_line(aes(sev, yl, linetype = "Yield loss"))+
  scale_linetype_manual(values = 2)+
  scale_color_colorblind()+
  scale_fill_colorblind()+
  scale_x_continuous(limits = c(0,100))+
  scale_y_continuous(limits = c(0,100),
                     # sec.axis = sec_axis(~ . * 1, name = "Yield loss (%)")
                     )+
  # facet_rep_wrap(~enso)+
  theme(strip.background = element_blank(),
        legend.position = "none")+
  coord_equal()+
  labs(x = "Severity (%)",
       y = "Soybean yield (%)",
       color = "ENSO phase",
       fill = "ENSO phase")
rel_gg

```



#### Combo plot

```{r}
 ((DC_gg/rel_gg) | forest_gg )+
  plot_annotation(tag_levels = "A")+
  # plot_layout(guides = "collect")&
  theme(legend.position = "bottom")
  
  
ggsave("figs/z.png", dpi = 900, width =7, height = 7.5, bg ="white")
```




# Yield Gain

> Vector for attainable yield (`ymax`), control efficacy (`lambda`), severity in untreated field (`sn`), and damage coefficient (`a`).

```{r}
ymax = seq(1500, 4000,length.out = 200)
lambda = c(30, 50, 70)
sn = seq(0, 1,length.out = 200)
a = dc_data$emmean
```

> Calculating yield gain (`yld_gain`) for the grid of vectors 


```{r}
ygain_df = expand.grid(ymax = ymax,lambda = lambda, sn = sn*100, a = a) %>% 
  mutate(yld_gain = (a*ymax*(sn - sn*((1-lambda/100))))/100) %>% 
  mutate(lambda = paste0(lambda,"% of Control")) %>% 
  full_join(dc_data %>% rename(a = emmean))
```


## Response surfaces

### Absolute yield gain

```{r}
ygain_df %>% 
  ggplot(aes(sn, ymax, fill = yld_gain))+ 
  geom_raster()+
  # scale_fill_distiller(palette = "Spectral")+
  # scale_fill_fermenter(palette = "Spectral")+
  scale_fill_viridis_b(option = "A", 
                       guide = guide_colorbar(barwidth = 15, barheight = 0.3),
                       breaks = seq(0, 3000, by =250)
                       )+
  facet_grid(lambda~enso)+
  theme_minimal_grid(font_size = 10)+
  labs(y = "Attainable yield (kg/ha)",
       x = "Severity untreated (%)",
       fill ="Yield gain (kg/ha)" )+
  theme(panel.grid = element_blank(),
        legend.position = "bottom")


ggsave("figs/surface_yield_gain.png", height =6, width = 6, bg = "white" )

```


### Yield gain difference from Neutral ENSO

```{r}
ygain_df %>% 
  group_by(enso) %>% 
  mutate(id = 1:length(enso)) %>% 
  ungroup() %>% 
  pivot_wider(id_cols = c(id,lambda, sn,ymax),
              values_from = yld_gain,
              names_from = enso) %>% 
  mutate(Cold = Cold - Neutral,
         Warm = Warm -Neutral) %>% 
  dplyr::select(-Neutral) %>% 
  pivot_longer(5:6,
               values_to = "diff",
               names_to = "enso") %>% 
  ggplot(aes(sn, ymax, fill = diff))+ 
  geom_raster()+
  scale_fill_steps2(midpoint = 0,
                    low = "#7b3014",
                    mid = "white",
                    high = "#27456e",
                    guide = guide_colorbar(barwidth = 13, barheight = 0.3),
                       breaks = seq(-300, 400, by =50)
                       )+
  scale_y_continuous(breaks = seq(min(ymax), max(ymax), length.out = 5))+
  scale_x_continuous(breaks = seq(min(sn)*100, max(sn)*100, length.out = 5 ))+
  facet_grid(lambda~enso)+
  theme_minimal_grid(font_size = 10)+
  labs(y = "Attainable yield (kg/ha)",
       x = "Severity untreated (%)",
       fill ="Yield gain difference from\nNeutral ENSO phase (kg/ha)" )+
  theme(panel.grid = element_blank(),
        legend.position = "bottom",
        axis.text = element_text(size =5),
        legend.text = element_text(size =5),
        legend.title = element_text(size =5))

ggsave("figs/surface_yield_gain_difference.png", height =5, width = 4, bg = "white" )
```


# Session Info

```{r}
sessionInfo()
```

